---
title: "R coding Exercise Gapminder, Tidyverse, and Plotting"
output: 
  html_document:
    toc: FALSE
---

First, we need to make sure all needed packages for the exercise are loaded using the command library(what we want to load), if not loaded (or you encounter an error of 'x' not found)- please install using the command install.packages(what you want to install)

For example, if dslabs is not installed you would use the command...install.packages("dslabs")

## PREPARATION

**Loads dslabs package**
```{r}
library("dslabs")
```

**Loads tidyverse (a collection of packages designed to help with cleaning data)**
```{r}
library('tidyverse')
```

**Loads scales package (to access break formatting functions)**
```{r}
library('scales')
```

<<<<<<< HEAD
**Loads hrbrthemes package (it provides additional themes, scales and components to make ggplot look nicer)**
```{r}
library('hrbrthemes')
```

**Loads viridis package (to access a wide range of color map that improves the graph presentation and readability)**
```{r}
library('viridis')
```

**Loads broom package (to convert output object from linear model to a table)**
```{r}
library('broom')
```
**Loads kableExtra package (we will need this package to create nice html table)**
```{r}
library("kableExtra")
```

=======
>>>>>>> main
**Looks at help file for the gapminder data**
```{r}
help(gapminder)
```


**Get an overview of the data structure**
```{r}
str(gapminder)
```


**Get a summary of the data**
```{r}
summary(gapminder)
```


**Determines the type of object gapminder is**
```{r}
class(gapminder)
```


## PART 1

**Filtering Data and creating New Variables**

We want to examine the data about African countries only, as a result, we will filter the data accordingly...

**Filter out data from all African countries (variable from continent= 'Africa') by using filter command, also use pipeline %<% to create standardized chain of commands (This will allow for easier processing aka "do this then that" )**

Note that the first part of this code (africadata<-gapminder) is needed due to the fact that we are assigning the African countries to a completely new object (africadata) for ease of analysis

**Creates new object and assigns all African countries to it**
```{r}
africadata <- gapminder %>% 
filter (continent == 'Africa')
```


Next we will recheck the structure and summary of the new data 

Note that since we are using the new object (africadata) we are no longer focusing the commands on gapminder

**Checks the structure and shows summary of the newly created object (africadata)**
```{r}
str(africadata)
summary(africadata)
```


Now that we have created a new object (africadata) and assigned all of the desired countries to it, we need to create new variables to help us analyze the data, in this example we will use the variables infant_mortality and Life_expectancy (which will be called INFMORT.LIFEEXP (variable 1)) and population and life_expectancy (which will be called POP.LIFEEXP (variable 2)).

**Note that we will use a similar command structure as when we assigned the gapminder data African countries to the new object africadata, only this time africadata will be replaced with our new desired objects (Variables 1 and 2 respectively)**

Please note that we also use the 'select' function as a way to choose which subset of variables we want to measure

We do not have to use the pipe function (%>%) since we are not doing several things at once (assigning and filtering in the previous case of africadata)

**Create another object (V1) in the africadata (object) with only infant mortality and life expectancy data**
```{r}
INFMORT.LIFEEXP <- select(africadata, infant_mortality, life_expectancy)
```


**Create another object (V2) in the africadata (object) with only population and life expectancy data**
```{r}
POP.LIFEEXP <- select(africadata, population, life_expectancy)
```



**We will now check the structure and summary of the two new objects we have created:**
```{r}
str(INFMORT.LIFEEXP)
summary(INFMORT.LIFEEXP)

str(POP.LIFEEXP)
summary(POP.LIFEEXP)
```


## PART 2

**Plotting**

For this section we will plot the new variables created- life expectancy as a function of infant mortality and as a function of population size. We will make two separate plots to compare this. 

**Note that since we are trying to create scatter plots (which means multiple commands) we will reuse the pipe function of %>%, making sure to use our new objects**

**To create a scatter plot the command function ggplot(aes(x,y)) is used, with geom_point() being added to signal the creation of a line (x and y lab simply label the x and y axis respectively)**

First Scatterplot of life expectancy as a function of infant mortality
```{r}
INFMORT.LIFEEXP %>% 
  ggplot(aes(x=infant_mortality, y=life_expectancy)) +
  geom_point() + 
  #add plot title
  ggtitle("Relationship between infant mortality and life expectancy for African countries\n from 1960 to 2016") +   
  #change x-axis label
  xlab('Infant Mortality (death per 1000)') + 
  #change y-axis label
  ylab('Life Expectancy (years)')
```


Second Scatterplot of life expectancy as a function of population 

**Note that here, the command scale_x_continous(trans = 'log10') is added along with the other commands. This is to set the x-axis to a log scale**
```{r}
POP.LIFEEXP %>% 
  ggplot(aes(x=population, y=life_expectancy))+
  geom_point() +
  #transform and reformat x-axis to log scale
  scale_x_continuous(trans = 'log10',
                    labels = trans_format("log10", math_format(10^.x))) +
  #add plot title
  ggtitle("Relationship between population size and life expectancy for African countries\n from 1960 to 2016") + 
  #change x-axis label
  xlab("Population (log scale)") + 
  #change y-axis label
  ylab("Life expectancy (years)") 
```

**Notice that right now, due to multiple observation years for each country being reported, the data is disorganized and has formed 'streaks'. To fix this, we must now narrow our data to a particular year.**

**We need to determine which years have missing data first to find the best year(s) for analysis. We can do this by utilizing the select function (to select the variable (year) and the variable that we are using the function for(infant_mortality)) combined with the filter command and the function (is.na), which is used to determine missing variables (na stands for non-applicable).**
```{r}
#write code that figures out which years have missing data for infant mortality
africadata %>% 
      group_by(year) %>% 
      summarise(infant_mortality_na = sum(is.na(infant_mortality))) %>% 
      filter(infant_mortality_na > 0) %>% 
      arrange(-infant_mortality_na)
```

  
  
**Now that we have determined that there are missing variables throughout the given years of the data set through 1981 and then again in 2016, we will go with the year 2000 for the most accurate analysis.**

Now we will extract only the year 2000 from the data
```{r}
africadata2000 <- africadata %>% 
  filter(year == 2000)
```


Now we will once again view the structure and summary to verify correct observations
```{r}
str(africadata2000)
summary(africadata2000)
```

Now that the data has been analyzed with only yeay 2000 being considered, we must now repeat the process done earlier to plot the newly separated data

**Please note that since the new data is assigned the object name of africadata2000, the object names of INFMORT.LIFEEXP and POP.LIFEEXP from the first scatter plots are both replaced by africadata2000**

**First scatter plot with new data only for year 2000**

```{r}
africadata2000 %>% 
  ggplot(aes(x=infant_mortality, y=life_expectancy)) +
  geom_point() + 
  #add plot title
  ggtitle("Relationship between infant mortality and life expectancy for African countries\n from 1960 to 2016") +   
  #change x-axis label
  xlab('Infant Mortality (death per 1000)') + 
  #change y-axis label
  ylab('Life Expectancy (years)')
```


**Second scattter plot with new data only for year 2000**
```{r}
africadata2000 %>% 
  ggplot(aes(x=population, y=life_expectancy))+
  geom_point() +
  #transform and reformat x-axis to log scale
  scale_x_continuous(trans = 'log10',
                    labels = trans_format("log10", math_format(10^.x))) +
  #add plot title
  ggtitle("Relationship between population size and life expectancy for African countries\n from 1960 to 2016") + 
  #change x-axis label
  xlab("Population (log scale)") + 
  #change y-axis label
  ylab("Life expectancy (years)") 
```

**The new scatter plots seem to suggest a negative correlation between infant mortality and life expectancy (still), while there is no evident correlation between population size and life expectancy.** 



## PART 3

**The last step we will need to do is to utilize linear regression models on both life expectancy and infant mortality, and life expectancy and population separately, to determine a possible link, if any.**

**To do this, we first establish the object names that we want to save the models to, which in this example will be 'fit1' and 'fit2'. Next, we utilize the lm function with the x-axis variable going second (infant mortality and population respectively). Finally, we utilize the summary function to gain an understanding of the models computations.**

```{r}
fit1 <- lm(life_expectancy ~ infant_mortality, data = africadata2000)

fit2 <- lm(life_expectancy ~ population, data = africadata2000)
```


**Summarize**
```{r}
summary(fit1)
summary(fit2)
```


Because of the final p-value of 2.826e-08 for the first model, which is less than 0.001, I can conclude that there exists a negative correlation between infant mortality and life expectancy, that is to say that for every unit of increase of infant mortality, life expectancy is predicted to decrease.

In the second model determining if there exists a link between population and life expectancy the p-value is greater than 0.005 at `r round(summary(fit2)$coefficient[2,4],4)`, which indicates that there is no statistically significant link between population size and life expectancy.


##  Additional Analysis by Tzu-Chun 
**Let's try to create a few more plots using data across all years**
**First, we will load a couple more packages for the tasks**

**Loads hrbrthemes package (it provides additional themes, scales and components to make ggplot look nicer)**
```{r}
library('hrbrthemes')
```

**Loads viridis package (to access a wide range of color map that improves the graph presentation and readability)**
```{r}
library('viridis')
```

**Loads broom package (to convert output object from lm to a table)**
```{r}
library('broom')
```
**Loads kableExtra package (we will need this package to create nice html table)**
```{r}
library("kableExtra")
```


**I would like to know if the distribution of life expectancy varies between different African regions. To accomplish that, we will create a boxplot with jitter to visualize the distribution of the life expectancy across different African regions.** 

```{r, warning = F}
africadata %>%
  #use fill to set a unique color for each region
  ggplot(aes(x=region, y=life_expectancy, fill=region)) +
  #add boxplot
  geom_boxplot() +
  #set color theme 
  scale_fill_viridis(discrete = TRUE, alpha=0.6, option="D") +
  #add jitter to plot
  geom_jitter(color="black", size=0.4, alpha=0.9) +
  #a clean and neat background to emphasize typography
  theme_ipsum() +
  #change title size and legend position
  theme(
    legend.position="none",
    plot.title = element_text(size=15)
    ) +
  #add plot title
  ggtitle("Comparison of life expectancy across five African regions") +
  #change x-axis label
  xlab("") + 
  #change y-axis label
  ylab("Life expectancy (years)") 
```

We can see that Northern Africa had the highest life expectancy but the data were also more spread out between countries. On the contrary, other four African regions had lower and similar life expectancy.

**Next, I want to know how GDP per capita across each country contributes to the change in life expectancy. The first step is to calculate the GDP per capita and create new variable by using mutate(). I then used features in ggplot2 package to create the scatter plot, but this time I will group the data by African region**
```{r}
#calculate GDP per capita and create new variable 
africadata <- africadata %>%
  mutate(gdp_per_capita = gdp/population)

africadata %>%
  #use color to set a unique color for each region
  ggplot(aes(x=gdp_per_capita, y=life_expectancy, color=factor(region))) +
  geom_point() +
  #change legend title
  guides(color=guide_legend(title="Region")) +
  #add plot title
  ggtitle("GDP per capita versus life expectancy across African regions\n from 1960 to 2016") + 
  #change x-axis label
  xlab("GDP per capita") + 
  #change y-axis label
  ylab("Life expectancy (years)")

```

The graph demonstrated a positive correlation between GDP per capita and life expectancy across all African regions. One more thing we noticed is that the rise in life expectancy by the unit change in GDP was higher among Northern and Eastern regions, however, the the similar trend was less apparent among other regions such as Middle or Southern Africa. 


**Now, we will fit a multiple linear regression models that used life expectancy as outcome and both infant mortality and GDP per capita as independent variables, to evaluate the strength of the relationship and output the statistics result in a table. I also fitted the same model again but added the interaction term.**

```{r}
#fit a multiple linear model to regress life expectancy on infant mortality and GDP per capita using full Africa data
fit3 <- lm(life_expectancy ~ infant_mortality + gdp_per_capita , data = africadata)
#summarize the lm output
summary(fit3)
```
```{r}
#use tidy to convert lm object to tibble
fit3 %>% 
  tidy() %>% 
  #round figures to show only 5 decimal places
  mutate_if(is.numeric, round, 5) %>% 
  #create an awesome HTML table
  kbl() %>% 
  kable_paper("hover", full_width = F)

```


```{r}
#fit a multiple linear model with interaction
fit4 <- lm(life_expectancy ~ infant_mortality*gdp_per_capita , data = africadata)
#summarize the lm output
summary(fit4)
```

```{r}
#use tidy to convert lm object to tibble
fit4 %>% 
  tidy() %>% 
  #round figures to show only 5 decimal places
  mutate_if(is.numeric, round, 5) %>% 
  #create an awesome HTML table
  kbl() %>% 
  kable_paper("hover", full_width = F)

```

The significant p-value (p<0.0001) for infant mortality indicated a linear relationship between infant mortality and life expectancy after adjusting for GDP per capita, and the coefficient showed that the life expectancy goes down by `r round(abs(summary(fit3)$coefficient[2,1]),2)` for every 1 unit increase in infant mortality. Similarly, we also found a linear relationship between GDP per capita and life expectancy after adjusting for infant mortality with a p-value smaller than 0.05. From the statistics output of the second model, we can see that when the slope GDP per capita increased, the slope for infant mortality decreased!
