---
title: "Tidy Tuesday"
output: 
  html_document:
    toc: FALSE
---

#Since this is our first time downloading a *tidytuesday* dataset, we must first 
#install the packages to allow for downloading of the data...
#Note that the data this week comes from the National Bureau of Economic Research with the 'nberwp' package

install.packages("tidytuesdayR")

#Now that we have the package installed we need to load the data for the week...

```{r}
papers <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-09-28/papers.csv')
authors <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-09-28/authors.csv')
programs <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-09-28/programs.csv')
paper_authors <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-09-28/paper_authors.csv')
paper_programs <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-09-28/paper_programs.csv')
```
#Next, we will utilize the cleaning script provided to combine the data...

```{r}
library(nberwp)
library(tidyverse)

papers %>% 
  write_csv("2021/2021-09-28/papers.csv")

authors %>% 
  write_csv("2021/2021-09-28/authors.csv")

programs %>% 
  write_csv("2021/2021-09-28/programs.csv")

paper_authors %>% 
  write_csv("2021/2021-09-28/paper_authors.csv")

paper_programs %>% 
  write_csv("2021/2021-09-28/paper_programs.csv")

joined_df <- left_join(papers, paper_authors) %>% 
  left_join(authors) %>% 
  left_join(paper_programs) %>% 
  left_join(programs)%>% 
  mutate(
    catalogue_group = str_sub(paper, 1, 1),
    catalogue_group = case_when(
      catalogue_group == "h" ~ "Historical",
      catalogue_group == "t" ~ "Technical",
      catalogue_group == "w" ~ "General"),
.after = paper) 

joined_df
```

#Now we will look at the structure of the data and examine it...

```{r}
str(joined_df)
```

#Now that we have organized the data, we will explore the data by creating a table
#that shows the number of each category of paper for each year...

```{r}
table1 <- table(joined_df$catalogue_group, joined_df$year)
table1
```

#We will create another table that shows the joined program and category 
#of each paper...
```{r}
table2 <- table(joined_df$program, joined_df$program_category)
table2
```

#Finally, we will make a third table that shows the programs with their 
#respective descriptions for further clarity...

```{r}
table3 <- table(joined_df$program, joined_df$program_desc)
table3
```

#Now we will perform a simple scatter plot with a line of best fit to determine 
#if there exists a pattern or trend between several different combinations of 
#factors

```{r}
Plot1 <- ggplot(joined_df,aes('year', 'program'))+
  geom_point(shape = 21, colour = "black", fill = "white", size = 5, stroke = 5)


Plot1
```